Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-07-21 manuscript/manuscript_file.Rmd
Most recent updates to manuscript

- include a better description of full range of lake size: sample and population.  also add NO3 to better support emission rate statement.
- incorporate in Methods section how/why size is incorporated as categorical variable and continuos variable.
- maybe discuss upper midwest.  Lots of lake area, but largely at equilibrium, therefore very small fluxes.  Large area doesn't necessarily mean big flux.
-  last sentence of emission rates.  we should look at summary stats of observed NO3 rather than predicted 
-  Mike will take a stab at big SI table.
-  Citation “epa_data” isn’t populating citation 26 in bibliography.  The bibTex entry is classified as a “website”.  Roy, can you attempt to make this citation play nice with nature_geosciences.csl?

## Introduction

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from lakes, reservoirs, and ponds, hereafter referred to as waterbodies, in the conterminous U.S. (CONUS). A generalized random-tessellation survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[weird, I have 1090, but NLA17 report says 1,005? compare my sample IDs to NLA IDs]** sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate (See NLA website for all data?). `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. We modeled the data using a Bayesion hierchical approach and predicted the dissolved N~2~O concentration for all 224,916 CONUS waterbodies in the target population **[where did I get 224,916, Roy has 465,897. See technical support document]**. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`), the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), and \>65% of the waterbodies in every ecoregion were undersaturated **[(Table 1)]**.

@@

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lentic waterbodies can function as an N~2~O sink. Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months and Soued et al. [@Soued_etal_2016] report that 40% of rivers, lakes, and ponds sampled during the summer in a boreal region of Quebec, Canada were functioning as N~2~O sinks **[29% of lakes in Finland (kortelainen et al 2020) but kortelainen and Soued are mostly pristine lakes]**. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

@@

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished.

@@

The model indicates that NO~3~, water temperature, and surface area interact to determine the N~2~O saturation ratio. These interactions yield a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enable the identifcation of broad scale patterns. First, small waterbodies with low NO~3~ concentration tend to be undersaturated, whereas small waterbodies with high NO~3~ tend to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, tend to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation can be well predicted by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

@@

**Not sure where I am going with this paragraph. Maybe scratch it. Mostly included as a placeholder for these references** While previous surveys have documented relationships between lake size and dissolved methane concentration (Pighini et al. 2018, Juutinen et al 2009, Borges et al. 2011), another greenhouse gas, to the best of our knowledge, this is the first study to document a relationship between lake size and $\Delta$ N~2~O. Perhaps because previous surveys have included little variation in waterbody size [@Webb_etal_2019; @Kortelainen_etal_2020] or water quality [@Soued_etal_2016].

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was `r national_stats$post_med_en2o` (95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**. At the ecoregional scale, only the Northern Appalachians and Northern Plains ecoregions had mean emission rates less than 0.  These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[need table or figure.  See ecoreg_stats$post_med_no3 variable.  maybe a SI table with summaries of all data from all_predictions?]**.

## Fluxes 
Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area.  The largest lake size class (> 50 Ha) had the greatest N~2~O flux in all ecoregions, but flux estimates for this lake size class were also the most uncertain **[(Fig. 6)]**. Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies.  Despite widespread N~2~O undersaturation, all ecoregions were a net source of N~2~O to the atmosphere **[(Table 1)]**.  Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface **[(Table 1)]**.  The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

@@

@@

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area.  The N~2~O flux from these lakes is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved N~2~O translate to large uncertainties in flux when propagated across the surface of large waterbodies.  Efforts to improve the accuracy of the lentic N~2~O flux estimate for the U.S. should therefore focus on investigations of the nations largest lakes and reservoirs.

The smallest lake size classes exhibited both the highest and lowest levels of N~2~O saturation, depending on NO~3~ availability.  It is unclear whether this variability is related to differences in NO~3~ loading or temporal variation.  For example, reservoirs have been shown to function as N~2~O sources in early summer when NO~3~ is replete, but switch to N~2~O sinks in later summer when the NO~3~ supply has been exhausted [@Beaulieu_etal_2014].  A better understanding of the temporal evolution of N~2~O production and consumption in small waterbodies is an important research need to determine the extent to which these systems can remove excess nitrogen from surface waters and simultaneously mitigate N~2~O emissions.

>>>>

- incorporate in Methods section how/why size is incorporated as categorical variable and continuos variable.
-  Mike will take a stab at big SI table.
-  Citation “epa_data” isn’t populating citation 11 in bibliography.  The bibTex entry is classified as a “website”.  Roy, can you attempt to make this citation play nice with nature_geosciences.csl?

## Psuedo abstract
Lakes, ponds, and reservoirs are thought to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018], but the relative lack of published N~2~O emission rates and recent evidence of N~2~O uptake across a broad range of waterbodies[@Soued_etal_2016;@Webb_etal_2019;@Kortelainen_etal_2020] have called into question the accuracy of emission estimates.   Here we measured dissolved N~2~O concentrations and calculated emission rates across `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[weird, I have 1090, but NLA17 report says 1,005? I get 984 if I exclude hand picked sites.  NLA17 site info file shows 1210 sites sampled.  I think the differences between these values are related to how hand picked and state intensification sites are handled.  The sampled we used contains data from 1090 sites, so that is what we should report here.]** waterbodies in the continental U.S. (CONUS).  Nitrous oxide undersaturation was widespread through the CONUS, with `r propN2oSinkMean` % of the waterbodies functioning as sinks for atmospheric N~2~O. Despite widespread N~2~O uptake by CONUS waterbodies, a small proportion of the waterbodies in each ecoregion had sufficiently high N~2~O emission rates to offset N~2~O uptake in the majority of waterbodies, such that all ecoregions functioned as net N~2~O sources to the atmosphere.  Waterbodies with high nitrate (NO~3~) tended to be sources of N~2~O whereas those with low NO~3~ were generally sinks.  Nitrate interacted with waterbody size, however, with large waterbodies having N~2~O concentrations closer to equilibrium than small waterbodies. We estimate that CONUS waterbodies emit `r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` )metric tons N~2~O year^-1^, equivalent to only 5% of N~2~O emissions from all CONUS surface waters (e.g. zero order streams through river mouths) per IPCC estimates .  The demonstrated capacity for widespread N~2~O uptake in CONUS waterbodies highlights the need to revisit global N~2~O models which presume surface waters are a N~2~O source [@epa_2023].

## Introduction
Lakes, reservoirs, and ponds (waterbodies hereafter) are considered globally significant sources of nitrous oxide (N~2~O)[@DelSontro_etal_2018], a potent greenhouse gas and contributor to stratospheric ozone destruction.  It is generally assumed that N~2~O emissions from surface waters scale with inorganic nitrogen loading [@Mosier_etal_1998] or productivity [@DelSontro_etal_2018], but recent reports of N~2~O uptake across waterbodies in Canada [@Soued_etal_2016;@Webb_etal_2019] and Finland [@Kortelainen_etal_2020] have called this assumption into question.  Furthermore, detailed investigations of individual waterbodies have shown that lentic N~2~O biogeochemistry is dynamic, with some waterbodies alternating between N~2~O sources and sinks in response to nitrogen availability, thermal stratification, and other factors [@Beaulieu_etal_2015;@Beaulieu_etal_2014;@Musenze_etal_2014].

While recent data sets demonstrate that waterbodies can function as N~2~O sinks, these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading).  Most work on aquatic N~2~O biogeochemistry has focused on streams and rivers, but these systems have been consistently found to be N~2~O sources and the controls on N~2~O biogeochmistry likely differ between flowing and standing waters.  These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. We modeled the data using a Bayesion hierchical approach and predicted the dissolved N~2~O concentration for all 465,897 CONUS waterbodies in the target population. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`), the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), and \>65% of the waterbodies in every ecoregion were undersaturated **[(Table 1)]**.

@@

The extent of N~2~O undersaturation in CONUS waterbodies is surprisingly high, particularly for waterbodies in nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lentic waterbodies can function as an N~2~O sink. Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months.  An investigation of boreal systems in Canada found that  40% of rivers, lakes, and ponds sampled during the summer were functioning as N~2~O sinks[@Soued_etal_2016]. Similarly, 29% of surveyed boreal lakes in Finland were found to be N~2~O sinks during the summer [@Kortelainen_etal_2020]. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lentic waterbodies function as N~2~O sinks during the summer months. This pattern holds across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use across the continental U.S..

@@

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished.  While previous surveys have documented relationships between lake size and dissolved methane concentration [@Pighini_etal_2018;@Juutinen_etal_2009;@Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

@@

@@

The model indicates that NO~3~ exerts a strong control on N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area.  This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be partially controlled by dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon and thermal stratification strength.  These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O.  Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020;@Yang_etal_2015] while others have found no relationship [@Soued_etal_2016;@Whitfield_etal_2011].  It is possible that the NO~3~ ~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

Interactions between NO~3~, water temperature, and lake size produced a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enable the identifcation of broad scale patterns. First, small waterbodies with low NO~3~ concentration tend to be undersaturated, whereas small waterbodies with high NO~3~ tend to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, tend to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation can be well predicted by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

@@

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which is expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015;@Kortelainen_etal_2020;@Webb_etal_2019].  Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was positive (`r national_stats$post_med_en2o` (95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^)) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**. At the ecoregional scale, only the Northern Appalachians and Northern Plains ecoregions had mean emission rates less than 0.  These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[(Table S1)]**.

## Fluxes 
Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area.  The largest lake size class (> 50 Ha) contained only `r round(1 - proportionSmallNumber, 2) * 100`% of the total number of lakes in the population, but had `r round(1-proportionSmallSA, 2) * 100`% of total lake surface area.  This lake size class had the greatest N~2~O flux, but flux estimates for this lake size class were also the most uncertain **[(Fig. 6)]**. Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies.  Despite widespread N~2~O undersaturation, all ecoregions were a net source of N~2~O to the atmosphere **[(Table 1)]**.  Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface **[(Table 1)]**.  The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

While total waterbody surface area contributed to differences in N~2~O flux among ecoregions, large cumulative surface area does not necessarily equate to large fluxes.  The Upper Midwest ecoregion had the second largest waterbody surface area, but only the sixth largest flux.  This is due to a mean saturation (mean = 1, 95% credible interview: 0.99 - 1.01) indicative of equilibirum with the atmosphere, which translates to very low fluxes.

@@

**[extract n2o data from Lauerwald for CONUS and compare]**

@@

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these waterbodies is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved N~2~O translate to large uncertainties in flux when propagated across the surface of large waterbodies. While our sample included some of the largest waterbodies in the CONUS **[(Table S2)]**, efforts to improve the accuracy of the lentic N~2~O flux estimate should include measurements at the nations largest lakes and reservoirs.

The smallest lake size classes exhibited both the highest and lowest levels of N~2~O saturation, depending on NO~3~ availability.  It is unclear whether this variability is related to differences in NO~3~ loading or is an artifact of temporal variation in NO~3~ processing and N~2~O production/consumption.  For example, reservoirs have been shown to function as N~2~O sources in early summer when NO~3~ is replete, but switch to N~2~O sinks in later summer when the NO~3~ supply has been exhausted [@Beaulieu_etal_2014].  Most small waterbodies in the CONUS function as N~2~O sinks during the summer, but a small subset are N~2~O emission hotspots.  A better understanding of the temporal evolution of N~2~O production and consumption is an important research need to determine the extent to which small waterbodies can remove excess nitrogen from surface waters and simultaneously mitigate N~2~O emissions.
